Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments

### 2021-05-14 doc/paper.Rmd
fix duplicate references

<!--# Insert page break -->

@@

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdam2004; @selwood2015].

@@

Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdam2004a; @selwood2015a; @laurance2001] are likely to be pervasive, but also far more complex than previously thought.

>>>>

\pagebreak

@@

Studies in temperate systems have shown that the demography of species can also be altered by climate change [@doak2010; @sletvold2005; @selwood2015; @williams2015], with more pronounced effects for fragmented populations [@oliver2015; @holyoak2016].
While the demographic consequences for tropical species are expected to be similarly severe [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], surprisingly little is known about the consequences of changing climates for taxa in tropical biomes, including Amazonia [@paniw2021].
Climate models for the next 50-100 years predict that some parts of the Amazon basin will experience more frequent and severe droughts as well as more frequent periods of high precipitation [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013; @duffy2015].
Plant populations may be particularly sensitive -- an increase in the frequency and severity of extreme precipitation events can have detrimental consequences for survival and reproduction [@estebanOtherSideDroughts2021; @gaoue2019].
This, coupled with the evidence that growth and survivorship are already lower in fragments [@lauranceRainForestFragmentation1998; @brunaExperimentalAssessmentHeliconia2002; @zartmanDroughtImpactsSurvivorship2015], has led to speculation that plants in tropical forest fragments are particularly at risk from drought and other climatic extremes [@lauranceEffectsStrongDrought2001; @opdamClimateChangeMeets2004; @selwood2015].

@@

Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We also found that the delayed responses of populations in fragments can differ significantly in magnitude, direction, and lag time from those of populations in continuous forest.
This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwood2015; @laurancePositiveFeedbacksForest2001] are likely to be pervasive, but also far more complex than previously thought.
